Constructing a universe for the setoid model

### 2020-07-06 paper.tex
small fixes

\[
\infer{
  \Prop : \Ty\ \GG
}{
  \GG : \Con
}
\qquad
\infer{
  \underline{P} : \Ty\ \GG
}{
  P : \Tm\ \GG\ \Prop
}
\]

@@

Another question has to do with the metatheoretic tools required to carry out
the construction of the universe. In fact, one of the main aspects of the setoid
model construction recalled in Section~\ref{setoid-model} and shown originally
in \cite{setoid99} is that it can be carried out in a very small type theoretic
metatheory, thus providing a way to reduce extensionality to a small intensional
core.
%
We would like to stay faithful to this ideal when constructing this setoid
universe.

@@

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support universe induction, thus exposing the inductive
structure of the codes. Such elimination principle is known to be inconsistent
with univalence, although this is not an issue in our case; nevertheless it's
not immediately clear if the elimination principle can be justified by the
semantics, that is, if our encoding of the setoid universe in the model allows
to define such universe eliminator.
%
The question arises because our final encoding of the setoid universe only
supports a weak form of elimination, for reasons that are explained in
Section~\ref{inductive-universe}. Although not currently needed, a stronger
eliminator might be necessary to justify universe induction. This problem should
not arise in the other encodings of the setoid universe (as given in
Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).

Another design choice has to do with how the setoid universe relates to the
other universes. One could have $\Prop$ be a code of the setoid universe, rather
than a type. Moreover, the setoid universes could form a hierarchy, possibly
cumulative.

@@

Induction-recursion is arguably a nicer and more natural way to define internal
universes in type theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$).

In our case, we can encode $\U$ as an inductive type $\ad{in-U}$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\ad{in-U}$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed on
the type of codes, the definition of $\ad{in-U}$ quite expectedly reflects the
inductive structure of codes.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \mType\} \{B : A \to \mType\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}

>>>>

\[
\infer{
  \Prop : \Ty\ \GG
}{
  \GG : \Con
}
\qquad
\infer{
  \underline{P} : \Ty\ \GG
}{
  P : \Tm\ \GG\ \Prop
}
\qquad
\infer{
  u \equiv v
}{
  u : \Tm\ \GG\ \underline{P} && v : \Tm\ \GG\ \underline{P}
}
\]

@@

Another question has to do with the metatheoretic tools required to carry out
the construction of the universe. In fact, one of the main aspects of the setoid
model construction recalled in Section~\ref{sec:setoid-model} and shown originally
in \cite{setoid99} is that it can be carried out in a very small type theoretic
metatheory, thus providing a way to reduce extensionality to a small intensional
core.
%
We would like to stay faithful to this ideal when constructing this setoid
universe.

@@

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support universe induction, thus exposing the inductive
structure of the codes. Such elimination principle is known to be inconsistent
with univalence, although this is not an issue in our case; nevertheless it's
not immediately clear if the elimination principle can be justified by the
semantics, that is, if our encoding of the setoid universe in the model allows
to define such a universe eliminator.
%
The question arises because our final encoding of the setoid universe only
supports a weak form of elimination, for reasons that are explained in
Section~\ref{inductive-universe}. Although not currently needed, a stronger
eliminator might be necessary to justify universe induction. This problem should
not arise in the other encodings of the setoid universe (as given in
Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).

Another design choice has to do with how the setoid universe relates to the
other universes. One could provide a code for $\Prop$ in the setoid universe.
Moreover, the setoid universes could form a hierarchy, possibly
cumulative.

@@

Induction-recursion is arguably a nicer and more natural way to define internal
universes in type theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$) \cite{malatasta13smallir}.

In our case, we can encode $\U$ as an inductive type $\ad{in-U}$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\ad{in-U}$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
the type of codes, the definition of $\ad{in-U}$ quite expectedly reflects the
inductive structure of codes.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \mType\} \{B : A \to \mType\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
